The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-11 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised Section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2549 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1f3bd10cffd26eae331ed6541b2a399ae9e918da

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e. the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Binarization of these residuals or scores.
(3)~Categorization of covariates.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

The classical CART algorithm \citep{Breiman+Friedman+Stone:1984} proposed to first grow a large
tree and then prune those splits in the tree that did not increase predictive performance in a
cross-validation. This is also know as \emph{postpruning} (after growing the initial tree) and
more specifically cost-complexity pruning.

In the unbiased recursive partitioning literature this postpruning approach is also used frequently
\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
from the association tests are only employed for selecting the split variable on a unified scale.
However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
so-called \emph{prepruning} strategy which stops growing the tree as soon as no significant association
can be found in a given subgroup. This approach is the default in CTree and also in
MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
postpruning in model-based partitioning is to use information criteria such as AIC (Akaike information
criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.

Clearly, for inference-based prepruning it is crucial that the association tests employed for
split selection work well as statistical significance test, i.e., conform with their nominal size
and have high power. In contrast, when using postpruning (either based on cross-validation or
information criteria) it might not be as crucial that the significance test works well and has
high power.

Due to these considerations we first evaluate the significance tests underlying CTree, MOB,
and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning strategy.
Subsequently we combine the tests with a cost-complexity postpruning approach in order to assess
whether shortcomings of the tests are mitigated by pruning.

\section{Unifying framework}\label{sec:unify}
Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of fitted models, the type of applied tests and the employed pruning strategy
as listed in Table~\ref{tab:combinations}. These settings represent the algorithms in their
original form. However, in principle all other combinations of the listed features are possible 
as well. As explained in Section~\ref{sec:pruning} the focus of this paper is on the testing 
strategies (second column of Table~\ref{tab:combinations}). They are here discussed for the case 
of parametric models together with prepruning as the properties of this pruning strategy
are already determined by the testing strategy. Subsequently, combinations
with postpruning are evaluated as well to show the impact of the choice of pruning strategy.
From here on the names CTree, MOB, and GUIDE are used to refer to the algorithm-specific
testing strategies rather than the complete algorithms.

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & Non-parametric  & Conditional inference    & Pre\\
MOB   & Parametric      & Score-based fluctuation  & Pre (or post with AIC/BIC)\\
GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
\hline
\end{tabular}
\caption{Combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.
\label{tab:combinations}}
\end{table}

\subsection{Unification of testing strategies}

Even though CTree, MOB, and GUIDE differ in the specific tests they apply (see second column of
Table~\ref{tab:combinations}) their approaches on how to select a split variable follow the same
basic structure as explained in Section~\ref{sec:generic_alg}. 
Thus, they can be presented together in a unifying framework for testing strategies
in model-based unbiased recursive partitioning. 
In this framework each specific testing strategy can be obtained by combining the provided 
building blocks accordingly. But apart from that, variations of the commonly known testing 
strategies can be obtained as this general toolbox allows for all possible combinations of the 
building blocks. Moreover, by exchanging only one of them its impact can be investigated in detail.
These building blocks are explained in Section~\ref{sec:buildingblocks} and their setting
for the standard setup of CTree, MOB, and GUIDE are listed in Table~\ref{tab:buildingblocks} which 
gives a more detailed overview of the differences in the second column of Table~\ref{tab:combinations}.

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}
The three following building blocks all refer to the way how the given data is prepared/transformed
before employing a statistical test to select a split variable.
\begin{itemize}
\item Variations of the goodness-of-fit measure:\\
Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
model that is obtained for the response $Y$ and possible covariates $X$ can be 
considered. This can be residuals $r(Y,X,\hat{\beta})$ as in the GUIDE algorithm or 
scores $s(Y,X,\hat{\beta})$ as in the MOB algorithm or, more generally, a 
transformation function $h(Y)$ as in the CTree algorithm including both of the foregoing cases.
By including all available scores rather than just residuals the goodness of fit of the 
model is investigated regarding each parameter instead of just the mean. 
%Therefore, this facilitates the detection of changes in parameters other than the mean/location parameter.

\item Binarization of residuals/scores:\\
Rather than applying the independence test directly on the residuals/scores a
binarized version of them can be used as well. In particular, the values are binarized
at $0$ such that only their signs are considered.

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd, 
and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$~test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

\end{itemize}

\textit{Note:}
Apart from these three building blocks, the algorithms CTree, MOB, and GUIDE differ in the type of test 
statistic they apply and their approximation of the underlying null distribution of the corresponding 
test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
and GUIDE ($\chi^2$~tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
introduced employing prepruning while GUIDE originally applies a postpruning strategy.
However, these differences refer to the standard versions of the three algorithms and all of them
allow for certain modifications of this standard setup. For example, a different form of test statistic
can be selected in CTree considering a maximum value as in MOB.

@@

\subsection{Linear model tree}\label{sec:lmtrees}
As an example a linear model tree is considered as this allows for an easy and well structured 
description of the investigated subjects. However, the conclusions drawn from this example also 
hold for a wide range of other model types. 
The models fitted in the nodes of the linear model tree as considered for the simulation 
study in Section~\ref{sec:simulation} are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
For each node the type of the fitted model is the same such that no node-specific features are 
considered and the observations of the (arbitrarily chosen) current node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of goodness-of-fit measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach one can consider residuals defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\ell$ is then
$$
\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding disaggregated, observation-wise goodness-of-fit measure is defined by
the score function 
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \ell}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\ell$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\ell$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

@@

In this simulation study two different scenarios are considered for the linear model trees
presented in Section~\ref{sec:lmtrees}. First, the underlying tree structure based on which the
data is generated is a stump, i.e. a tree with only one split. 
By keeping the complexity of the model as low as possible the advantages and
disadvantages of the different testing strategies can be elaborated more precisely and 
the focus can be put on their power in terms of selecting the correct split variable.

@@

\subsection{Data generating process}
Each generated data set consists of a response variable, one regressor, one split 
variable and nine noise variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

@@

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
(can) depend on the true split variable $Z_1$. 
%in the following way:
%\begin{center}
%$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
%\end{center}
%
In particular, three different cases are considered for the coefficients $\beta_0$ 
and $\beta_1$:
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
(at $0$).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}

The varying coefficient(s) $\beta_i$, $i\in\{0,1\}$, are binary taking either the positive or
the negative value of the effect size, i.e. $\delta$ or $-\delta$.
% can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
% In the binary case the following definition is applied for the true split point $\xi$
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{i}  \quad \text{if } Z_1 < \xi \\
+\delta \cdot (-1)^{i}  \quad \text{if } Z_1 \geq \xi
\end{cases}
\end{align*}
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.
An illustration of the three different cases 
%for binary coefficients and a continuous regressor 
can be found in Figure~\ref{fig:dgp_stump}.

@@

As explained before, the aim of this simulation study is to investigate the effects
of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
rather than the whole testing strategies. Therefore, other combinations as presented in 
Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
as their original versions. Moreover, as prepruning is employed all of the considered statistical 
tests are applied as significance tests answering two questions at once: 
(1)~Should a split be performed at all? 
(2)~If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.

@@

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e. only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
only scores has already been elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
in Section~\ref{sec:discussion}.

@@

Altogether these results clearly illustrate that overall the decision to include all available
scores when testing for dependencies in order to select split variables is advantageous over
considering only residuals, i.e. the scores regarding only the intercept, as changes not only in
the mean but also in all other parameters can be detected in that way. This is particularly obvious
when comparing GUIDE and GUIDE\_scores. Moreover, from the fact that these two version are both
performing worse than MOB and CTree it can be deduced that categorizing split variables and
binarizing scores has a negative effect on the ability to detect the right split variable. 
To elaborate these effects more in detail the results for all possible combinations of the 
building blocks are discussed in a full factorial 3-way analysis.

\subsection{Full factorial analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e. $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
in order to make the differences particularly well observable.
In the right panel the true split point is close to the upper end of the considered interval
($\xi=0.8$) requiring a higher effect size $\delta = 1$ in order to visualize results properly.
(Additional evaluations for increasing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

To elaborate the impact of the choice of a pruning strategy
the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
applying postpruning (building very large trees without any significance level and employing 
cost-complexity pruning as in the CART algorithm \citep{Breiman+Friedman+Stone:1984}). 
In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with prepruning,
again based on the Adjusted Rand Index.

For MOB and CTree it has been shown that the tests work very well as significance tests.
Therefore, it can be recommended to take advantage of this power and use the information
given by the tests for prepruning. As it can be seen in Figure~\ref{fig:pruning_ari}
in most cases the results of these two testing strategies do not vary significantly over 
changing the applied pruning strategy.
On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
strategy should not be based on their results which argues for the application of postpruning.
This can also be observed in Figure~\ref{fig:pruning_ari} showing that postpruning improves
the results for both GUIDE versions.

@@

pr <- simres$res
## extract data for xyplot comparing ARI and ARI_p
pr <- pr[,c(1,2,7,8,9,12,15)]
## extended data set with postpruning results being represented as separated tests
prex <- pr[,c(1,2,3,4,6)]
prex <- rbind(prex,prex)

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(2,2,2,2,1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings
)
@
\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 data sets of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). 
The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE\_scores. 
The dashed lines represent the results from applying prepruning and the continuous lines 
the results from applying postpruning.}
\end{figure}

>>>>

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Binarization of these residuals or scores.
(3)~Categorization of covariates.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

The classical CART algorithm \citep{Breiman+Friedman+Stone:1984} proposed to first grow a large
tree and then prune those splits in the tree that did not increase predictive performance in a
cross-validation. This is also know as \emph{post-pruning} (after growing the initial tree) and
more specifically cost-complexity pruning.

In the unbiased recursive partitioning literature this post-pruning approach is also used frequently
\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
from the association tests are only employed for selecting the split variable on a unified scale.
However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
so-called \emph{pre-pruning} strategy which stops growing the tree as soon as no significant association
can be found in a given subgroup. This approach is the default in CTree and also in
MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
post-pruning in model-based partitioning is to use information criteria such as AIC (Akaike information
criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.

Clearly, for inference-based pre-pruning it is crucial that the association tests employed for
split selection work well as statistical significance test, i.e., conform with their nominal size
and have high power. In contrast, when using post-pruning (either based on cross-validation or
information criteria) it might not be as crucial that the significance test works well and has
high power.

Due to these considerations we first evaluate the significance tests underlying CTree, MOB,
and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning strategy.
Subsequently we combine the tests with a cost-complexity post-pruning approach in order to assess
whether shortcomings of the tests are mitigated by pruning.

\section{Unifying framework}\label{sec:unify}

Each of the algorithms CTree, GUIDE, and MOB can be characterized by its combination
of the type of model fits, tests, and pruning strategy employed to
grow the tree. Table~\ref{tab:combinations} provides an overview of the default
combinations.
However, as discussed above, subsequent publications have emphasized that all three
algorithms can be combined with different model fitting approaches and to some degree
different pruning strategies have been explored as well. Thus, the class of tests
employed for the unbiased splitting variable selection forms the core of each of the
algorithms: conditional inference (CTree) vs.\ score-based fluctuation tests (MOB)
vs.\ residual-based $\chi^2$ tests (GUIDE). Therefore, we consider a standard class
of model fits (namely, linear regression trees) and investigate the relative
advantages and disadvantages of the tests themselves (which are essential to pre-pruning)
as well as the combination with post-pruning. Thus, subsequently
the names CTree, MOB, and GUIDE distinguish the test-based variable and split selection
rather than the entire algorithm with all default settings.

\begin{table}[t!]
\centering
\begin{tabular}{ l l l l }
\hline 
      & Fit             & Test                     & Pruning \\ 
\hline
CTree & Non-parametric  & Conditional inference    & Pre\\
MOB   & Parametric      & Score-based fluctuation  & Pre (or post with AIC/BIC)\\
GUIDE & Parametric      & Residual-based $\chi^2$  & Post (cost-complexity pruning)\\
\hline
\end{tabular}
\caption{Default combinations of fitted model type, test type and pruning strategy
for the algorithms CTree, MOB, and GUIDE.
\label{tab:combinations}}
\end{table}

\subsection{Building blocks of testing strategies}\label{sec:buildingblocks}

Even though CTree, MOB, and GUIDE differ in the specific tests they apply, their approaches for split
variable selection follow the same basic structure as explained in Section~\ref{sec:generic_alg}. 
In fact, the tests can be embedded in a unifying conceptual framework that yields the
different tests by combining various building blocks.
These mostly differ in the way the model for the dependent variable on the one hand and
the splitting variables on the other are prepared or transformed:
%
\begin{itemize}
  \item \emph{Goodness-of-fit measure:}\\
        Different variations of the disaggregated, observation-wise goodness-of-fit measure of the 
        model for the response $Y$ and possible covariates $X$ can be considered. Either residuals
	$r(Y,X,\hat{\beta})$ can be used as proposed for GUIDE or model scores $s(Y,X,\hat{\beta})$
        as proposed for MOB. All three algorithms can, in principle, use both goodness-of-fit
	measures though, which is probably brought out most clearly in the CTree algorithm that
        explicitly allowed for different transformations $h(Y)$ in its original description already.
  \item \emph{Binarization of residuals/scores:}\\
        Rather than testing independence between the split variables and the residuals/scores
	tmeselves, it is possible	
	to binarize residuals/scores at $0$ so that only their signs are assessed (as proposed for GUIDE).
  \item \emph{Categorization of split variables:}\\
        Similarly, the split variables can also be categorized (for testing only). This was proposed for
	GUIDE, employing binning at the quartiles yielding four categories of approximately
	equal size.
\end{itemize}
%
The three algorithms combine these building blocks in different ways as shown in Table~\ref{tab:buildingblocks}:
When applying CTree for unbiased model-based recursive partitioning it has been suggested to use
the model scores without binarization and assess their association with the untransformed split
variables using a conditional inference test. This is similar to a squared correlation test statistic.
MOB also employs the scores without
binarization and maximally selects a score statistic over all potential split points in the
split variable. GUIDE employs the binarized residuals and assesses their association with
the categorized split variable in a classical (unconditional) $\chi^2$ test.

But the building blocks could be easily re-combined to yield new types of tests. For example,
in the GUIDE approach, a 1-way ANOVA can be used for assessing the association of the residuals
(without binarization) with the categorized split variable. Or alternatively, a multivariate
1-way ANOVA can be used for the non-binarized scores as opposed to the residuals etc.

Note that there are further differences in the testing strategies between the three algorithms, e.g.,
using conditional vs.\ unconditional approximations of the null distributions. However, this difference
has relatively little influence compared to the other building blocks considered in detail here.
Moreover, both similarities and relative differences between these approaches have been previously
discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.

@@

\subsection{Linear model tree}\label{sec:lmtrees}

To focus on the unified testing framework, as described in the previous section, we employ
the same model fits for all three algorithms. To do so, we employ linear regression models
because it is such a basic and widely used model and linear model trees were the leading
illustrations in both the original MOB \citep{Zeileis+Hothorn+Hornik:2008} and GUIDE \citep{Loh:2002} papers.
However, the conclusions drawn from this example also hold for many other model types.

To fix notation, we consider the following models for the simulation 
study in Section~\ref{sec:simulation}:
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
with response variable $Y$, regressor variable $X$, and error term $\epsilon$. This model is
fitted, as usual, by ordinary least squares (OLS) to the observations in each notation,
yielding the parameter estimates $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
To keep notation simple we present the following equations for the root node with all observations
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ and $n \in \mathbb{N}$. In subsequent nodes the same equations
are used but just for a smaller subgroup.

The aggregated goodness-of-fit measure (or objective or loss function) in OLS estimation
is the sum of squared residuals: $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ where
$$
\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
is the squared residual which is defined as
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
These residuals can also intuitively be used as the corresponding disaggregated
observation-wise goodness-of-fit measure. Another natural candidate for this is the
score function for the $i$-th observation:
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \ell}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
Thus, up to a constant scaling factor of $-2$ (that could also be omitted) the first component
of the scores is in fact the residual. However, it is complemented by a second component
that captures the slope effect of $x_i$. Therefore, when computing the score for all $n$ observations
this yields an $n \times 2$ score matrix whose first column corresponds to the $n$ residuals.
As this is the derivative with respect to the intercept parameter~$\beta_0$, it
captures changes in the intercept. Moreover, the second column of the score matrix contains
derivatives with respect to the slope parameter~$\beta_1$ and thus captures changes in this.

@@

In this simulation study two different scenarios are considered for the linear model trees
presented in Section~\ref{sec:lmtrees}. First, the underlying tree structure based on which the
data is generated is a stump, i.e., a tree with only one split. 
By keeping the complexity of the model as low as possible the advantages and
disadvantages of the different testing strategies can be elaborated in more detail and 
focus can be given to their power in terms of selecting the correct split variable.

@@

\subsection{Data generating process}
Each generated data set consists of the response and regressor variables, one true split 
variable and nine noise split variables as listed in Table~\ref{tab:variables} 
together with the corresponding distributions.

@@

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$ of this linear equation
(can) depend on the true split variable $Z_1$. More specifically, three different scenarios
are considered for the coefficients $\beta_0$ and $\beta_1$:
%
\begin{enumerate}
  \item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
  \item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
        (at $0$).
  \item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
The varying coefficient(s) $\beta_j$, $j\in\{0,1\}$, are binary taking either the positive or
the negative value of the effect size, i.e., $\delta$ or $-\delta$.
% can either be binary (taking either $\delta$ or $-\delta$), or a continuous function. 
% In the binary case the following definition is applied for the true split point $\xi$
\begin{align*}
\beta_{i}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{j}  \quad \text{if } Z_1 < \xi \\
+\delta \cdot (-1)^{j}  \quad \text{if } Z_1 \geq \xi
\end{cases}
\end{align*}
%while in the continuous case a simple linear function is considered in the form of
%$$\beta_{i}(Z_1) = Z_1 \cdot \delta \cdot (-1)^{i}.$$
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.
An illustration of the three different cases 
%for binary coefficients and a continuous regressor 
can be found in Figure~\ref{fig:dgp_stump}.

@@

As explained before, the aim of this simulation study is to investigate the effects
of each particular building block of the unifying framework presented in Section~\ref{sec:unify} 
rather than the whole testing strategies. Therefore, other combinations as presented in 
Table~\ref{tab:buildingblocks}, hence adapted versions of GUIDE, CTREE, and MOB are evaluated as well
as their original versions. Moreover, as pre-pruning is employed all of the considered statistical 
tests are applied as significance tests answering two questions at once: 
(1)~Should a split be performed at all? 
(2)~If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha = 0.05$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is selected.

@@

For a central split point ($\xi=0$, upper panels) all strategies perform similarly if the 
intercept varies (left and right panels) where CTree and MOB are only slightly ahead. However, 
this advantage increases if the slope parameter varies and the intercept is kept at a fixed value 
(middle panel). In particular, the original strategy of GUIDE fails to detect any changes 
in $\beta_1$. 
This is due to the fact that GUIDE only includes residuals, i.e., only the scores 
regarding the intercept, such that changes in the slope parameter are not detected unless the 
corresponding scores are also included as in GUIDE\_scores. This disadvantage of using 
only scores has already been elaborated in the literature of structural change
analysis \citep[see e.g.][]{Ploberger+Kramer:1992} and is also discussed 
in Section~\ref{sec:discussion}.

@@

Altogether these results clearly illustrate that overall the decision to include all available
scores when testing for dependencies in order to select split variables is advantageous over
considering only residuals, i.e., the scores regarding only the intercept, as changes not only in
the mean but also in all other parameters can be detected in that way. This is particularly obvious
when comparing GUIDE and GUIDE\_scores. Moreover, from the fact that these two version are both
performing worse than MOB and CTree it can be deduced that categorizing split variables and
binarizing scores has a negative effect on the ability to detect the right split variable. 
To elaborate these effects more in detail the results for all possible combinations of the 
building blocks are discussed in a full factorial 3-way analysis.

\subsection{Full factorial analysis}\label{sec:3way}
In Figure~\ref{fig:3way} the effects of the building blocks are illustrated for the most general
case where the intercept and slope parameter are both varying. In the left panel the true split point is in the center, i.e., $\xi=0$ and an effect size of $\delta = 0.3$ is considered. 
This choice has been made based on the results shown in Figure~\ref{fig:scores_stump} 
in order to make the differences particularly well observable.
In the right panel the true split point is close to the upper end of the considered interval
($\xi=0.8$) requiring a higher effect size $\delta = 1$ in order to visualize results properly.
(Additional evaluations for increasing $\delta$ as in Section~\ref{sec:scores} 
can be found in Appendix~\ref{app:increasingdelta}).
The power of the different combinations is measured by the mean $p$-value for the true split 
variable~$Z_1$.
The possible levels for each of the three building blocks are listed in Table~\ref{tab:levels}.
In total 12 combinations are applied including the original testing strategies CTree, MOB, and GUIDE
and variations of them. A more detailed overview of all 12 combinations can be found in 
Appendix~\ref{app:combinations}, Table~\ref{tab:combinations_appendix}.

@@

To elaborate the impact of the choice of a pruning strategy
the tree models presented in Figure~\ref{fig:tree_ari} are built again, but this time
applying post-pruning (building very large trees without any significance level and employing 
cost-complexity pruning as in the CART algorithm \citep{Breiman+Friedman+Stone:1984}). 
In Figure~\ref{fig:pruning_ari} the results are compared to those obtained with pre-pruning,
again based on the Adjusted Rand Index.

For MOB and CTree it has been shown that the tests work very well as significance tests.
Therefore, it can be recommended to take advantage of this power and use the information
given by the tests for pre-pruning. As it can be seen in Figure~\ref{fig:pruning_ari}
in most cases the results of these two testing strategies do not vary significantly over 
changing the applied pruning strategy.
On the contrary, overall the tests in GUIDE have less power. For that reason, the pruning
strategy should not be based on their results which argues for the application of post-pruning.
This can also be observed in Figure~\ref{fig:pruning_ari} showing that post-pruning improves
the results for both GUIDE versions.

@@

pr <- simres$res
## extract data for xyplot comparing ARI and ARI_p
pr <- pr[,c(1,2,7,8,9,12,15)]
## extended data set with post-pruning results being represented as separated tests
prex <- pr[,c(1,2,3,4,6)]
prex <- rbind(prex,prex)

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<prune_ari, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(ari ~ delta | xi, groups = ~ test, 
       data = prex, 
       type = "l", lty = c(2,2,2,2,1,1,1,1),
       lwd = 2,
       key =  list(text = list(levels(prex$test)[1:4]),
                   lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                col = pal[c("CTree", "MOB", "GUIDE_scores", "GUIDE")], lwd = 2)),
       layout= c(4,1),
       scales=list(x=list(at=seq(1,11,2))),
       xlab = expression(delta),
       ylab = "Adjusted Rand Index",
       strip = strip.custom(factor.levels = c(expression(xi==0), 
                                              expression(xi==0.2), 
                                              expression(xi==0.5), 
                                              expression(xi==0.8)),
                                   bg = "gray90"),
       par.settings = par.settings
)
@
\caption{\label{fig:pruning_ari}Adjusted Rand Index over increasing effect size $\delta$ 
(100 data sets of 250 observations per step).
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right). 
The performances of four testing strategies are compared: CTree, MOB, GUIDE, GUIDE\_scores. 
The dashed lines represent the results from applying pre-pruning and the continuous lines 
the results from applying post-pruning.}
\end{figure}
